---
title: "Using proteasy to Retrieve and Analyze Protease Data"
author: "Martin Rydén"
package: proteasy
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Using proteasy to Retrieve and Analyze Protease Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteKeywords{Proteomics, Bioinformatics,
  Cleavage, Polypeptides, MEROPS, Proteases}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{proteasy}
---

```{r biocstyle, echo = FALSE, results = "asis", message = FALSE}
library(BiocStyle)
BiocStyle::markdown()

```

# Introduction

This package utilizes data derived from the
[MEROPS database](https://www.ebi.ac.uk/merops/).
The database is a manually curated knowledgebase with information about
proteolytic enzymes (proteases), their inhibitors and substrates. 

Similarly to existing tools such as
[TopFind](https://topfind.clip.msl.ubc.ca/), and
[Proteasix](http://www.proteasix.org), `r BiocStyle::Biocpkg("proteasy")`
exists for the purpose of retrieving data on proteolytic enzymes by mapping
peptide termini to known sites where a protease cleaves. Unlike the
`r BiocStyle::Biocpkg("cleaver")` package, which allows for in-silico cleavage
of amino acid sequences by specifying an enzyme, the functions in
`r BiocStyle::Biocpkg("proteasy")` relies only on experimentally derived data
to find proteases annotated to cleavage sites.

This document illustrates the functionality of proteasy through some use cases.

# Installation

The package should be installed using as described below:

```{r installPackage, eval = FALSE}

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("proteasy")

```

The package can then be loaded with:

```{r loadPackage, eval = TRUE}

library("proteasy")

```

# Usage

## Find known proteases by substrate search

A fast way to find which proteases, if any, are annotated as cleaving actors
for a substrate is by using the function *searchSubstrate*. Setting the 
parameter summarize = TRUE will return only a vector of reviewed proteases,
and summarize = FALSE will return a table with details about each cleaving
event. We will explore the two options for kininogen-1 (P01042).

```{r SearchSubstrate, eval = TRUE}

# Returns vector of reviewed proteases that cleave P01042
searchSubstrate(protein = "P01042", summarize = TRUE)

# Returns data.table with details on cleaving events
searchSubstrate(protein = "P01042", summarize = FALSE) |>
    head()

# The function also accepts multiple proteins as input. Let's inspect
# the last rows in the returned data.table.
searchSubstrate(protein = c("P01042", "P02461"), summarize = FALSE) |>
    tail()

# With summarize = FALSE we get both reviewed and unreviewed proteases.

```


## Find substrates by protease search

A corresponding function, *searchProtease*, exists to find which (if any)
substrates a protease cleaves. The function is demonstrated with
MMP-12 (P39900) as an example.

```{r SearchProtease, eval = TRUE}

# Returns vector of substrates that MMP-12 cleave
searchProtease(protein = "P39900", summarize = TRUE)

# Returns data.table with details on cleaving events that involve MMP-12
searchProtease(protein = "P39900", summarize = FALSE) |> head()

```

## Find possible proteases for cleaved peptides

The function *findProtease* automatically maps the peptide sequences to the
full-length protein sequence and obtains the start- and end-positions for the
peptide. Then, the positions are searched against the MEROPs database and
matches are returned.


```{r FindProtease, eval = TRUE}

# Create a vector of Uniprot IDs and corresponding peptide sequences

protein <- c("P02671", "P02671", "P68871",
             "P01011", "P68133", "P02461",
             "P0DJI8", "P0DJI8", "P0DJI8")
peptide <- c("FEEVSGNVSPGTR", "FVSETESR", "LLVVYPW",
             "ITLLSAL", "DSYVGDEAQS", "AGGFAPYYG",
             "FFSFLGEAFDGAR", "EANYIGSDKY", "GGVWAAEAISDAR")

# If we do not specify start position (start_pos) and end position
# (end_pos), the function will automatically assign these values by
# matching the provided peptide sequence against the full-length
# protein sequence.

res <- findProtease(protein = protein,
                    peptide = peptide,
                    organism = "Homo sapiens")

# The resulting S4 object can be inspected in three ways;
# (to save space we show only the first six rows)

# 1. Display sequence data for the provided input:

substrates(res) |> head()

# 2. Show which known proteases may have cleaved this protein:

proteases(res) |> head()

# 3. Display details of associated proteolytic events:

cleavages(res) |> head()

# We can find out what proportion of matching cleaving events by reviewed
# proteases occur at N- versus C-terminus

cl <- cleavages(res)[`Protease status` == "reviewed"]

cl$`Cleaved terminus` |> table() |> prop.table() |> round(digits = 2)

# And inspect the distribution of cleaved amino acids

cl$`Cleaved residue` |> table() |> barplot(cex.names = 2)


# Find which protease is involved in the greatest number of cleaving events

cl[!duplicated(Peptide), .(count = .N), by = `Protease (Uniprot)`]

# If start- and end-positions are specified, the function will not attempt
# to automatically look up sequence data for the specified protein/peptide.

cl_by_pos <- findProtease(
    protein = "P02671",
    peptide = "FEEVSGNVSPGTR",
    start_pos = 413,
    end_pos = 425
)

# However, this means sequence details for substrates is not available.

substrates(cl_by_pos)



```

## Look up a protease on MEROPS

We may want to read up on the details of an entry directly in MEROPS.
The function *browseProtease* takes a UniProt or MEROPS ID and opens the 
MEROPS summary page which corresponds to that ID in a web browser.

```{r Browse protease, eval = FALSE}

browseProtease("P07339", keytype = "UniprotID") # (opens web browser)

```

# Additional examples

## Plot cleavages as a protein-protein interaction network.

```{r network, eval = TRUE, message = FALSE, fig.height = 12, fig.width = 12}

library(igraph)
library(data.table)

# Make a vector of substrates we want to investigate
proteins <- c('P01011','P02671')

# Look up known proteases which cleave these substrates, and associated data
# annotated to the cleaving events
res <- searchSubstrate(protein = proteins, summarize = FALSE)

# Filter to keep proteases with Uniprot status "reviewed"
res <- res[`Protease status` == "reviewed"]

# To create a network, we only need two columns of interactors
# i.e. the proteases with cleaving action on the substrates
res <- res[, c("Protease (Uniprot)", "Substrate (Uniprot)", "Cleavage type")]

# Construct the DAG
g <- igraph::graph_from_data_frame(res,
                                   directed = TRUE,
                                   vertices = unique(
                                       c(res$`Protease (Uniprot)`,
                                         res$`Substrate (Uniprot)`)))

# This will allow us to return to a layout we like
# (and where the legend fits nicely)
set.seed(104)

plot(g,
     vertex.label.family = "Helvetica",
     vertex.size = 14,
     vertex.color = ifelse(V(g)$name %in% res$`Protease (Uniprot)`,
                           "#377EB8", "#E41A1C"),
     vertex.label.cex = 1,
     vertex.label.color = "white",
     edge.arrow.size = .6,
     edge.color =  ifelse(E(g)$`Cleavage type` == "physiological",
                          "#01665E", "#8E0152"),
     layout = igraph::layout.davidson.harel)

legend(title = "Nodes", cex = 2, horiz = FALSE,
       title.adj = 0.0, inset = c(0.0, 0.2),
       "bottomleft", bty = "n",
       legend = c("Protease", "Substrate"),
       fill = c("#377EB8", "#E41A1C"), border = NA)

legend(title = "Edges", cex = 2, horiz = FALSE,
       title.adj = 0.0, inset = c(0.0, 0.0),
       "bottomleft", bty = "n",
       legend = c("Physiological", "Non-physiological"),
       fill = c("#01665E", "#8E0152"), border = NA)

```

# Session Information

```{r sessioninfo, echo = FALSE}
sessionInfo()
```
